Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-16 paper.tex
Updates from duncan and Yan

It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
and affects the results of downstream analysis, e.g., inference of
selection~\citep{Anisimova2003-vr}. However, standard phylogenetic methods do
not account for recombination
\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Instead,
genetic ancestries of recombining lineages can be represented by ancestral
recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-properties}, which
capture network-like phylogenetic relationships \citep{Gusfield2014-qw}. Recently,
development of the ``succinct tree sequence'' data
structure \citep{Kelleher2018-xc}  has enabled compact yet flexible
encoding of large ARGs. This representation should be well-suited to storing
large recombinant viral genealogies such as SARS-CoV-2,
which currently encompasses over 15 million sequences in the GISAID database
\citep{Shu2017-hp}). Moreover the \emph{tskit} software library, which implements this
data structure, enables efficient manipulation of the stored genealogies and is
integrated with a suite of open-source software tools to simulate, analyse, and
infer them \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.

@@

Here, we present a preliminary version of a novel scalable methodology
(\emph{sc2ts}, optionally pronounced ``scoots'') that builds on the advances above
and additionally takes advantage of known collection dates to infer ARGs from millions
of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2 genome sequences,
we demonstrate that \emph{sc2ts} can be used to simultaneously (1) detect putative SARS-CoV-2
recombinants; (2) infer the phylogenetic relationships of the major SARS-CoV-2 lineages;
(3) reconstruct the mutational spectrum of SARS-CoV-2; and (4) provide a compact record
of evolutionary information about SARS-CoV-2 that can be efficiently accessed using
\emph{tskit} (tree sequence toolkit) software library \citep{Kelleher2018-xc}.
\emph{Sc2ts} is a promising method to automate detection of SARS-CoV-2 recombinants
and to jointly model recombinational and mutational processes shaping the genomic
diversity of SARS-CoV-2 at scale.

@@

Although recombination nodes are relatively infrequent, the wide and long ARGs still
contain many of them (4123 and 2078 respectively; Supplementary Tables 1 and 2).
Each of these nodes were
inferred as a result of matching a specific recombinant sample into the ARG.
However, ~63\% of recombination nodes are ancestral to only a single
recombinant sample; we exclude these "singleton recombinants" from subsequent
analyses as they may have arisen from sequencing of mixed samples (e.g., from
co-infection or contamination), and in any case, never result in onward
transmission.  This leaves 1522 and 763 recombination nodes in the wide and
long ARGs, respectively. In their analysis using RIPPLES,
\cite{Turakhia2022-it} detected 589 non-singleton recombinants.

@@

As a check on the reliability of these identified recombinations,
Figure~\ref{fig:recomb_mrcas}B  plots only those breakpoints which are
HMM-consistent. This reduces both the number of recombination nodes \&
breakpoints in the long ARG from 763 \& 851 to 347 \& 358
(and in the wide ARG from 1522 \& 1769 to XXX \& YYY , see
Supplementary Tables 1 and 2).

The date of common ancestry for the two parents of recombinant lineages is
concentrated in several banded rows in Figure~\ref{fig:recomb_mrcas}. These are
largely due to a small handful of MRCAs which are shared between a large number
of recombination events; the top four are indicated by dotted lines in the
figure. These shared MRCAs are near the root of large expansions of certain
lineages (B.1: original variant, B.1.617.2: Delta, BA.1: first Omicron wave,
and BA.2: second Omicron wave), as evidenced by the number of samples with those
Pango designations which trace back to each ancestral node (listed as
percentages on the plot). The majority of these frequently-shared MRCAs
are associated with extensive polytomies in the tree (at the extreme,
the BA.2-linked node has 2,872 immediate children). This is likely to
indicate a rapid and under-sampled expansion of a clade in the SARS-CoV-2 genealogy.

@@

The remaining remaining 630 Pango recombinant samples, representing 31
Nextclade-designated Pango lineages, trace back to 50 different recombination nodes.
We examine the inferred inheritance structure in the following subsections.

@@

\caption{\label{fig:pango-simple-origin-graph} Simple origin
Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
imputed Pango status. (B) Equivalent plot for the 17 XAG samples in the long ARG, with abbreviated labelling.
(C) Equivalent plot for both origination events involving the four XD samples in the long ARG. Details of the mutations and sample node identities for all three plots are provided in supplementary figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}.
} \end{figure}

A simple example is the case of XA, illustrated in Figure~\ref{fig:pango-simple-origin-graph}A.
This shows a subgraph of the long ARG, in which paths have been traced from all
Nextclade-identified XA samples (in red) to the closest other sample nodes in our dataset. The
subgraph may thus include inserted, non-sample nodes (i.e. UPGMA, recombination, mutation
collapse, or reversion push nodes, all plotted as smaller circles) on the path to the nearest
sample nodes (larger circles). Dotted lines show where this subgraph links to the rest of the
ARG. Above recombination nodes only ancestral nodes are shown, so that e.g. cousins of non-sample
recombinant parents are not shown. It is clear from the subgraph that all the XA samples trace
to a single originating
recombination node (plotted in black), with a breakpoint position around 22227bp.

@@

A more complex single-origin case is XAG, illustrated in
Figure~\ref{fig:pango-simple-origin-graph}B. Here, the XAG samples
all trace back to the same most recent common recombination node (black), but
we infer this recombination event to also be the originating event for
all the recombinant samples designated XAA,
and some, but not all, of those identified as XAB, XQ, and XU by Nextclade.

The classification of originating recombination events is dependent on
accurate designation of Pango lineages to samples. It is therefore important
to note that if the GISAID Pango designations are used, many of the samples
marked here as XAB are reclassified
as BA.2 (and XAG becomes fully monophyletic, see supplementary figure
\ref{fig:pango_XAG_gisaid_graph}).

In total we identify 16/31 single-origin Pango recombinant lineages in the long ARG.
Supplementary Table~\ref{tab:pango-single-origin} summarises these origination events
together with their inferred parents, showing good agreement between \texttt{sc2ts} and
the official Pango designated parents, with \texttt{sc2ts} providing more a precise
parent designation. Six of the 16 lineages (XA, XAC, XAE, XF, XK, and XS) are of the
simple-origin (XA) type, whereas the rest are of the XAG type, and include multiple Pango
designated lineages as descendants of the originating recombination event. In some cases,
these may be caused by samples which have been designated as the wrong recombinant lineage
by Nextclade.

\subsubsection{Multiple origins}
The ARG inference has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. In this section we examine the Pango X lineage
samples that are inferred to have multiple recombinant origins (i.e. trace
back to independent recombination nodes in the ARG).

Figure~\ref{fig:pango-simple-origin-graph}C shows a simple example, consisting of
the 4 samples labelled XD by Nextclade in the long ARG. The left hand subgraph
(the origin of three XD samples, all sampled in France) has an earliest
sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
subgraph has a single sample with EPI\_ISL 8514045 (sampled in Turkey and dated 2021-12-30).
The different geographical locations, the time between the samples, and
the fact that the two samples differ at 23 nucleotide positions,
suggests that these may indeed represent true independent origins of XD.
Supplementary Figure~\ref{fig:pango_XD_gisaid_graph} shows the same subgraphs
with sample identifiers and coloured by GISAID Pango labels (however, no
samples are designated by GISAID as XD in the long ARG).

Another X lineage we infer to have multiple origins is XE. In this case, 
we identify at least 4 independent origination events of the Nextclade-designated
XE samples (Supplementary Table~\ref{tab:multiple_origins_table}).

[TODO: needs more detail]

@@

\begin{figure} \centering
\includegraphics[width=0.5\textwidth]{figures/ls_diagram_covid.pdf}
% FIXME this caption is too long
\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS)
model, in which a query sequence (bottom) is described as an
imperfect mosaic of the sequences in a reference panel.
Black crosses along the focal sequence show genotyping
errors or mutations.
In the standard formulation, at site $\ell$, the recombination probability is $r_\ell$,
the mutation probability is $\mu_\ell$ and $n$
denotes the size of the reference panel.
The Viterbi algorithm can be used to find a
``copying path'' through the reference panel for a given query sequence that
maximises the likelihood under these parameters. Unseen states in the reference panel are shown as coloured lines enclosed by
the grey box. The black arrow describes the true path through the data which leads to the emitted
focal sequence below. Examples of transition and
emission probabilities along this trajectory are shown by the red and blue
arrows, respectively.
}
\end{figure}

@@

The generative process of the LS model is summarised in
Figure~\ref{fig:ls_diagram}. Here, a transition matrix, $Q$, governs the
process of switching (recombining) between members of the reference panel (the
hidden states). An emission matrix, $E$, allows for differences between the
focal sequence and the hidden state from which it is copied (due to mutation as
well as genotyping or sequencing error). Both $E$ and $Q$ may be a function of the reference panel members,
but transitions are generally assumed to be independent of the hidden states (Figure~\ref{fig:ls_diagram}, pink panel).
This assumption dramatically increases performance as the state space drops to two states (i.e., switching or not switching).
Emissions may also be a function of the nucleotide states, but in our
case, we assume that all observed alleles (in both the reference panel and
focal sequence, $a\ell$) at the site under consideration are equally likely and
occur with mutation probability $\frac{\mu_\ell}{a_\ell-1}$ (Figure~\ref{fig:ls_diagram}, blue panel).
We use the Viterbi algorithm
\citep{Viterbi1967-ol} to find the most likely copying path through the
sequences, given a set of reference sequences, $Q$, and $E$.
The relative rates of mutation and recombination are usually controlled by
the site-specific parameters $\mu_\ell$ and $r_\ell$, respectively.
For this work we used a slightly different formulation, which uses a parameter, the mismatch ratio (MMR), to control the relative importance of mutation
and recombination in the HMM. Specifically, an MMR value of $k$ will
prefer $k$ mismatches (mutations) to a single recombination which results in copying from a template with no mismatches. To map between recombination probability and mutation probability for a particular mismatch ratio, we simply consider the two paths which we wish to be equally likely, and rearrange for the mutation or recombination probability. In the simple case when we assume $\mu_l=\mu$ and $r_l=r$, we obtain (supplementary methods):
\begin{align*}
r = \frac{n\mu^k}{(1-\mu)^k + (n-1)\mu^k}, \quad \quad
\mu = \frac{1}{\sqrt[k]{\frac{n}{r} - (n-1)} + 1}.
\end{align*}
Under a lower MMR value (for which recombinations are favoured over mutations), the maximum-likelihood HMM copying path of a
focal genome more likely spans at least two ancestral genomes (i.e,. the focal
genome is a putative recombinant) than a single ancestral genome. Here, the MMR
was set to three in all the HMM runs.

A varying degree of missingness is present in the genetic data, that is, for a given sample, a subset of nucleotides coded as `unknown' at a set of positions. As a result, we had to consider missingness in our implementation of LS Viterbi. One approach would be to `impute' missing data, and hard-call or define some distribution over the possible missing nucleotide, and use that in the determination of path probabilities. Another is to not attempt to estimate the true nucleotide, but rather to consider such sites as uninformative to the path probability. We chose the latter. To do this, we simply encode the emission probability where the underlying nucleotide transition is $\mathit{unknown}\rightarrow \{A,C,G,T\}$, $\{A,C,G,T\}\rightarrow\mathit{unknown}$, or $\mathit{unknown}\rightarrow\mathit{unknown}$ at 1.

% [ALSO MENTION MISSING DATA, where we deal with Ns in the alignment]
% OLD text, ready to be adapted (and deleted once done)
% For this
% application, we have parameterised the HMM using a mismatch ratio (MMR), which
% tunes how strongly mutations (i.e., no switching of template sequences) are
% preferred over recombination (i.e., switching between sequences) during
% copying.

In \texttt{sc2ts} we use the efficient ARG-based implementation of the
LS Viterbi algorithm from \texttt{tsinfer} \citep{Kelleher2019-ba} to find
the most likely copying path for each sample sequence
among all sequences (sampled and inferred) in the current ARG.
In the majority of cases, with non-recombinant sample sequences,
the most likely solution is to copy from one
of the nodes in the ARG that minimised the number of mutations required
on the query sequence. Importantly, because the reference panel here consists of
\emph{every} node in the ARG, we can match to both older sample
sequences or internal nodes representing an inferred ancestral sequence
(see subsequent subsections for details about how these are added).
Thus, when no recombination is present, the LS Viterbi algorithm is
implementing a version of parsimony, in which we are guaranteed to
find a sequence that minimises the number of additional mutations required
to incorporate a newly-added sample into the ARG.
Recombination is then inferred when the most likely solution to the LS
HMM is to copy from more than one ARG node along the genome, for a
given sample sequence. [TODO MORE]

@@

It is important to note that the Viterbi algorithm only returns
\emph{one of} the copying paths that maximise the likelihood under
 the given mutation and recombination parameters. There may be many
such paths, from which we choose one arbitrarily. Also, the
present choice of using a single mismatch-ratio parameter to control
the likelihood of recombination vs mutation may lead to relatively
flat likelihood spaces where many different paths have equal likelihood.
There are many possibilities in using established HMM methodology
to reason about and explore the space of possible matches, which may
be a fruitful avenue for future work. See the Discussion for further
potential improvements that may be made to the HMM.

@@

\renewcommand\thefigure{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand\thetable{S\arabic{table}}
\setcounter{table}{0}
\section*{Supplementary}
\subsection*{Supplementary methods}
\subsubsection*{Derivation of relation between recombination probability and mismatch probability conditional on mismatch ratio}
Without loss of generality, consider a region of length $m$. Up until this region, two paths are equally likely, so we can re-scale by the likelihood up to the site before the region starts, $c$.

For a MMR of $k$, we only need to consider two of the $n$ members of the reference panel. One; $\rho_1$, for which there are no mismatches, but we need to recombine in, and a second; $\rho_2$, for which there are $k$ (randomly chosen) mismatches in the region.

We can simply follow the probability of tracking along each of those paths:
\begin{align*}
P(\rho_1) &= c \frac{r}{n} \left(\left(1-r\right)+\frac{r}{n}\right)^m\left(1-\mu\right)^m &\text{recombine to a template with no mismatches,}\\
P(\rho_2) &= c \left(\left(1-r\right) + \frac{r}{n}\right)\left(\left(1-r\right)+\frac{r}{n}\right)^m \left(1-\mu\right)^{m-k}\mu^k &\text{stay in a template with $k$ mismatches,}
\end{align*}
where $r$ is the recombination probability between adjacent variants, $\mu$ is the mismatch probability between a target sequence at a variant and a particular member of the reference panel, and $c$ is the likelihood of the path up to this point, which we assume to be equal.
We then set these path probabilities equal, and rearrange to obtain:
\begin{align*}
r &= \frac{n\mu^k}{\left(1-\mu\right)^k + \left(n-1\right)\mu^k},\\
\mu &= \frac{1}{\sqrt[k]{\frac{n}{r} - (n-1)} + 1}
\end{align*}

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/supp_cophylogeny_long.pdf}
\caption{\label{fig:cophylogeny_long}Tanglegram equivalent to that in Figure~\ref{fig:cophylogeny}, but for the long ARG (i.e. subsampled to mid-2022). }
\end{figure}

>>>>

It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
and affects the results of downstream analysis, e.g., inference of
selection~\citep{Anisimova2003-vr}. However, standard phylogenetic methods do
not account for recombination
\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Instead,
genetic ancestries of recombining lineages can be represented by ancestral
recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-properties}, which
capture network-like phylogenetic relationships \citep{Gusfield2014-qw}. Recently,
development of the ``succinct tree sequence'' data
structure \citep{Kelleher2018-xc}  has enabled compact yet flexible
encoding of large ARGs. This representation should be well-suited to storing
large recombinant viral genealogies such as SARS-CoV-2,
which currently encompasses over 15 million sequences in the GISAID database
\citep{Shu2017-hp}). Moreover the \texttt{tskit} software library, which implements this
data structure, enables efficient manipulation of the stored genealogies and is
integrated with a suite of open-source software tools to simulate, analyse, and
infer them \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.

@@

Here, we present a preliminary version of a novel scalable methodology
(\texttt{sc2ts}, optionally pronounced ``scoots'') that builds on the advances above
and additionally takes advantage of known collection dates to infer ARGs from millions
of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2 genome sequences,
we demonstrate that \texttt{sc2ts} can be used to simultaneously (1) detect putative SARS-CoV-2
recombinants; (2) infer the phylogenetic relationships of the major SARS-CoV-2 lineages;
(3) reconstruct the mutational spectrum of SARS-CoV-2; and (4) provide a compact record
of evolutionary information about SARS-CoV-2 that can be efficiently accessed using
\texttt{tskit} (tree sequence toolkit) software library \citep{Kelleher2018-xc}.
\texttt{Sc2ts} is a promising method to automate detection of SARS-CoV-2 recombinants
and to jointly model recombinational and mutational processes shaping the genomic
diversity of SARS-CoV-2 at scale.

@@

Although recombination nodes are relatively infrequent, there are still many
present in the wide and long ARGs (4123 and 2078 respectively; Supplementary
Tables 1 and 2). Each recombination node was inferred as a
result of matching a specific (``causal'') recombinant sample into the ARG.
However, ~63\% of them are ancestral to only a single
recombinant sample; we exclude these "singleton recombinants" from subsequent
analyses as they may have arisen from sequencing of mixed samples (e.g., from
co-infection or contamination), and in any case, never result in onward
transmission.  This leaves 1522 and 763 recombination nodes in the wide and
long ARGs, respectively. In their analysis using RIPPLES,
\cite{Turakhia2022-it} detected 589 non-singleton recombinants.

@@

As a check on the reliability of these identified recombinations,
Figure~\ref{fig:recomb_mrcas}B  plots only breakpoints which are
HMM-consistent. This reduces both the number of recombination nodes \&
breakpoints in the long ARG from 763 \& 851 to 347 \& 358
(and in the wide ARG from 1522 \& 1769 to XXX \& YYY , see
Supplementary Tables 1 and 2).

The date of common ancestry for the two parents of recombinant lineages is
concentrated in several banded rows in Figure~\ref{fig:recomb_mrcas}. These are
largely due to a small handful of MRCAs which are shared between a large number
of recombination events; the top four are indicated by dotted lines in the
figure. These shared MRCAs lie near the root of large expansions of certain
lineages (B.1: original variant, B.1.617.2: Delta, BA.1: first Omicron wave,
and BA.2: second Omicron wave), as evidenced by the number of samples with those
Pango designations which trace back to each ancestral node (listed as
percentages on the plot). The majority of these frequently-shared MRCAs
are associated with extensive polytomies in the tree (at the extreme,
the BA.2-linked node has 2,872 immediate children). This is likely to
indicate a rapid and under-sampled expansion of a clade in the SARS-CoV-2 genealogy.

@@

The remaining remaining 630 Pango recombinant samples, representing 31
Nextclade-designated Pango lineages, trace back to 50 different most recent
recombination nodes.
We examine the inferred inheritance structure in the following subsections.

@@

\caption{\label{fig:pango-simple-origin-graph} Simple origin
Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
imputed Pango status. Genomic regions inherited by the recombination node use 
breakpoint positions in the ARG, corresponding to the maximum inferred breakpoint position
(B) Equivalent plot for the 17 XAG samples in the long ARG, with abbreviated labelling.
(C) Equivalent plot for both origination events involving the four XD samples in the long ARG. Details of the mutations and sample node identities for all three plots are provided in supplementary figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}.
} \end{figure}

A simple example is the case of XA, illustrated in Figure~\ref{fig:pango-simple-origin-graph}A.
This shows a subgraph of the long ARG, in which paths have been traced from all
Nextclade-identified XA samples (in red) to the closest other sample nodes in our dataset.
Sample nodes are plotted as larger circles, but the subgraph also includes intermediate,
non-sample nodes: i.e. inserted nodes such as UPGMA, recombination, mutation collapse, and
reversion push nodes, plotted as smaller circles. Dotted lines show where this subgraph links
to the rest of the ARG. Above recombination nodes, only ancestral nodes are shown, meaning
that the subgraph is not extended to show other descendants of recombinant parents. It is clear
from the subgraph that all the XA
samples trace to a single originating recombination node,
whose genome is a composite of a B.1.177.18 lineage on the left of the genome (up to a maximum
position of ${\sim}22227$ basepairs) and a B.1.1.7 lineage for the remainder of the genome.

@@

A more complex single-origin case is XAG, illustrated in
Figure~\ref{fig:pango-simple-origin-graph}B. Here, the XAG samples
all trace back to the same most recent common recombination node 
(combining BA.1 on the left and BA.2.9 on the right), but
we infer this recombination event to also be the originating event for
all the recombinant samples designated XAA,
and some, but not all, of those identified as XAB, XQ, and XU by Nextclade.

The classification of originating recombination events is dependent on
accurate designation of Pango lineages to samples. It is therefore important
to note that if the GISAID Pango designations are used, many of the samples
marked here as XAB are reclassified as BA.2 (and XAG becomes fully monophyletic,
see supplementary figure \ref{fig:pango_XAG_gisaid_graph}). This is an
independent confirmation of the uncertainty in designation of these XAG-related
samples.

In total out of the 31 Pango recombinant lineages in the long ARG we identify 16 as
having a single origin. Supplementary Table~\ref{tab:pango-single-origin} summarises these
together with their inferred parents, showing good agreement between \texttt{sc2ts} and
the official Pango designated parents, with \texttt{sc2ts} providing more a precise
parent designation. Six of the 16 lineages (XA, XAC, XAE, XF, XK, and XS) are of the
simple-origin (XA) type, whereas the rest are of the XAG type, and include multiple Pango
designated lineages as descendants of the originating recombination event. In some cases,
these may be, however, be a result of erroneous Pango designations.

\subsubsection{Multiple origins}
The ARG inference has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. In this section we examine several Pango X lineage
samples that are inferred to have multiple recombinant origins (i.e. trace
back to independent recombination nodes in the ARG).

Figure~\ref{fig:pango-simple-origin-graph}C shows a simple example, consisting of
the 4 samples labelled XD by Nextclade in the long ARG. The left hand subgraph
(the origin of three XD samples, all sampled in France) has an earliest
sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
subgraph has a single sample with EPI\_ISL~8514045 (sampled in Turkey and dated 2021-12-30).
Both involve an Omicron lineage being inserted into the middle of a Delta genome, but
the breakpoints in each case are slightly different: the start of the Omicron insertion
in the French samples has an estimated maximum position of 21641 basepairs
(and minimum of 21619) with the insertion end occurring at a maximum position of 25584
(and a minimum of 25470). In contrast, the Omicron insertion in the Turkish sample
is inferred to have occurred between positions 22578 (min 21619) and 24130 (min 23605).
The breakpoint difference, the different geographical locations, the time between the
samples, and the fact that the two samples differ at 23 nucleotide positions,
suggests that these may indeed represent true independent Deltacron recombinants.
The canonical XD definition is based entirely on samples from northern Europe, particularly
France (\url{https://github.com/cov-lineages/pango-designation/issues/444}) so it seems
plausible that the earlier Turkish sample has been mislabelled XD by Nextclade.
Indeed, GISAID does not label any of these samples XD (see Supplementary
Figure~\ref{fig:pango_XD_gisaid_graph} which gives exact mutations and sample identifiers).

A similar pattern is seen for several of the origins inferred for the Pango
XE lineage, which was originally defined on the basis of sequences from the UK and Ireland. The
Long ARG infers five independent origins for nonsingleton Nextclade-designated XE
samples (Supplementary Table~\ref{tab:multiple_origins_table}). One of the recombinant origins,
caused in our ARG by an English sample, accounts for 155 out of 164 XE samples
(primarily UK and Irish, but with a few northern European and north American strains).
This certainly represents the canonical XE recombinant lineage. The other recombination nodes
only give rise for one, two, or three XE samples. Three of these recombination
nodes are likely to represent Nextclade mislabelling: they either have different
inferred breakpoint positions or (in the case of the earliest recombination node),
descendants that are primarily labelled as other Pango X-lineages. Additional support
for the suggestion that they result from incorrect labelling comes from the fact that
none of their XE-labelled descendant samples are from the UK or Ireland.
However, the most recently derived XE cluster, consisting
of two samples (one from Canada and one from England) may not be due to mislabelling: not
only does it contain an English sample but it has exactly the same breakpoint interval as
the canonical XE clade. It is possible that instead, this is a case where the \texttt{sc2ts} algorithm
has taken a "true" XE sample and mistakenly inferred an additional, spurious origin. [WHY? - this
is presumably because it has changed too much from the nearest in the main XE cluster - need to discuss the
weighting of 3 mutations for one recombination here. NB: the number
of nucleotide differences between the closest XE in the two clusters is 11: see the
Pango-recombinant-analysis workbook]

@@

\begin{figure} \centering
\includegraphics[width=0.5\textwidth]{figures/ls_diagram_covid.pdf}
\caption{\label{fig:ls_diagram} A schematic of the Li and Stephens (LS)
model, in which a query sequence (bottom) is described as an
imperfect mosaic of the sequences in a reference panel.
Black crosses along the focal sequence show genotyping/sequencing
errors or mutations.
In the standard formulation, at site $\ell$, the recombination probability is $r_\ell$,
the mutation probability is $\mu_\ell$ and $n$
denotes the size of the reference panel.
The Viterbi algorithm can be used to find a
``copying path'' through the reference panel for a given query sequence that
maximises the likelihood under these parameters. Unseen states in the reference panel are shown as coloured lines enclosed by
the grey box. The black arrow describes the true path through the data which leads to the emitted
focal sequence below. Examples of transition and
emission probabilities along this trajectory are shown by the red and blue
arrows, respectively.
}
\end{figure}

@@

The generative process of the LS model is summarised in
Figure~\ref{fig:ls_diagram}. Here, a transition matrix, $Q$, governs the
process of switching (recombining) between members of the reference panel (the
hidden states). An emission matrix, $E$, allows for differences between the
focal sequence and the hidden state from which it is copied (due to mutation as
well as genotyping or sequencing error). Both $E$ and $Q$ may be a function of the reference panel members,
but transitions are generally assumed to be independent of the hidden states (Figure~\ref{fig:ls_diagram}, pink panel).
This assumption dramatically increases performance as the state space drops to two states (i.e., switching or not switching).
Emissions may also be a function of the nucleotide states, but in our
case, we assume that all observed alleles (in both the reference panel and
focal sequence, $a_\ell$) at the site under consideration are equally likely and
occur with mutation probability $\frac{\mu_\ell}{a_\ell-1}$ (Figure~\ref{fig:ls_diagram}, blue panel).
We use the Viterbi algorithm
\citep{Viterbi1967-ol} to find the most likely copying path through the
sequences, given a set of reference sequences, $Q$, and $E$. Throughout, we refer to the probabilities of mismatching to a member of the reference panel at site $\ell$; $\mu_\ell$,
and the probability of recombining between two members of the reference panel between site $\ell-1$ and site $\ell$; $r_\ell$). $\mu_\ell$ is often referred to as the 
`mutation probability' for convenience, but we note that this probability of 
mismatch also encompasses various other error modes which result in a mismatch
(such as sequencing and alignment errors). Note also that it is these probabilities,
not the rates of mutation and recombination, that are required to fully define the HMM.
That said, we can approximately relate these probabilities to the coalescent with recombination process which LS is approximating. In the case of recombination,
suppose that there is some constant underlying recombination rate, $\rho$.
Assuming that recombination events occur as a Poisson process, 
then the probability of recombining between adjacent sites is:
\begin{align*}
    r_\ell = 1 - \exp\left(-\frac{\rho\left(m_{i-1} - m_{i}\right)}{n}\right),
\end{align*}
where $m_j$ is the nucleotide position of variant $j$. Similarly, we may consider the probability of mutation to represent to the 
probability of mutation along the branch joining the focal sequence to a coalescent
tree containing a reference panel with $n$ members, assuming a Poisson process
along the branch. The expected length of this branch is $\frac{4}{n}$, and so the mismatch 
probability represents the probability of observing a mutation or error mode, 
along this branch. We can then approximate the probability of observing a mismatch as:
\begin{align}
    \mu_\ell \approx 1 - \exp{\left(-\frac{4\nu}{n}\right)}
\end{align}
where $\nu$ is the mutation rate per $\mathit{ploidy} \times N_e$ generations. The probabilities of these competing process of mismatch and recombination are usually controlled by the site-specific parameters $\mu_\ell$ and $r_\ell$, respectively.
For this work we used a slightly different formulation, which uses a parameter, the mismatch ratio (MMR), to control the relative importance of mutation
and recombination in the HMM. Specifically, an MMR value of $k$ will
prefer $k$ mismatches (mutations) to a single recombination which results in copying from a template with no mismatches. 
To map between recombination probability and mutation probability for a particular mismatch
ratio, we simply consider the two paths which we wish to be equally likely, and rearrange 
for the mutation or recombination probability. Consider the simple case when we assume 
$\mu_l=\mu$ and $r_l=r$. Without loss of generality, consider a region of length $m$. Up 
until this region, two paths are equally likely, so we can re-scale by the likelihood of 
observing the focal sequence up to the site before the region starts, $c$. For a MMR of $k$,
we only need to consider two of the $n$ members of the reference panel. One; 
$\mathcal{P}_1$, for which there are no mismatches, but we need to recombine in, and a 
second; $\mathcal{P}_2$, for which there are $k$ (randomly chosen) mismatches in the region.
Now, follow the probability of tracking along each of those paths:
\begin{align*}
P(\mathcal{P}_1) &= \frac{cr}{n} \left(\left(1-r\right)+\frac{r}{n}\right)^m\left(1-\mu\right)^m &\text{recombine to a template with no mismatches,}\\
P(\mathcal{P}_2) &= c \left(\left(1-r\right) + \frac{r}{n}\right)\left(\left(1-r\right)+\frac{r}{n}\right)^m \left(1-\mu\right)^{m-k}\mu^k &\text{stay in a template with $k$ mismatches,}
\end{align*}
where $r$ is the recombination probability between adjacent variants, $\mu$ is the mismatch
probability between a target sequence at a variant and a particular member of the reference
panel, and $c$ is the likelihood of the path up to this point, which we assume to be equal.
We then set these path probabilities equal, and rearrange to obtain:
\begin{align*}
r = \frac{n\mu^k}{\left(1-\mu\right)^k + \left(n-1\right)\mu^k}, \quad\quad
\mu = \frac{1}{\sqrt[k]{\frac{n}{r} - (n-1)} + 1}.
\end{align*}
Under a lower MMR value (for which recombinations are favoured over mutations), the maximum-likelihood HMM copying path of a
focal genome more likely spans at least two ancestral genomes (i.e,. the focal
genome is a putative recombinant) than a single ancestral genome. Here, the MMR
was set to three in all the HMM runs.

A varying degree of missingness is present in the genetic data, that is, for a given sample, 
a subset of nucleotides coded as `unknown' at a set of positions. As a result, we had to
consider missingness in our implementation of LS Viterbi. One approach would be to `impute'
all missing data beforehand using external software to hard-call or define a 
distribution over the possible missing nucleotides, and use that in the determination of path
probabilities. Another is to not attempt to estimate the true nucleotide, but rather to 
consider such sites as uninformative to the path probability, but to impute missing data 
within that path on the fly, placing them within the reference panel iteratively. We chose the latter. To do this, we simply encode the emission
probability where the underlying nucleotide transition is $\mathit{unknown}\rightarrow \{A,C,G,T\}$ as 1.

In \texttt{sc2ts} we use the efficient ARG-based implementation of the
LS Viterbi algorithm from \texttt{tsinfer} \citep{Kelleher2019-ba} to find
the most likely copying path for each sample sequence
among all sequences (sampled and inferred) in the current ARG.
In the majority of cases, with non-recombinant sample sequences,
the most likely solution is to copy from one
of the nodes in the ARG that minimised the number of mutations required
on the query sequence. Importantly, because the reference panel here consists of
\emph{every} node in the ARG, we can match to both older sample
sequences or internal nodes representing an inferred ancestral sequence
(see subsequent subsections for details about how these are added).
Thus, when no recombination is present, the LS Viterbi algorithm is
implementing a version of parsimony, in which we are guaranteed to
find a sequence that minimises the number of additional mutations required
to incorporate a newly-added sample into the ARG.
Recombination is then inferred when the most likely solution to the LS
HMM is to copy from more than one ARG node along the genome, for a
given sample sequence. This statement warrants emphasis: the Viterbi algorithm enables us to find a path through the reference panel from among the $n^m$ paths that is provably at the optimum, under the LS model.

@@

It is important to note that the Viterbi algorithm only returns
\emph{one of} the copying paths that maximise the likelihood under
 the given mutation and recombination parameters. There may be many
such paths, from which we choose one arbitrarily. Also, the
present choice of using a single mismatch-ratio parameter to control
the likelihood of recombination vs mutation may lead to relatively
flat likelihood spaces where many different paths have equal likelihood.
There are many possibilities in using established HMM methodology
to reason about and explore the space of possible matches, which may
be a fruitful avenue for future work. Examples include a stochastic traceback through the collection of paths at the global optimum to glean further information about our the likelihood surface, and determine whether there are downstream implications for our conclusions. Here, we have considered the Viterbi algorithm to make statements about the most likely paths through the data. The machinery used here can be modified to run the forwards and backwards algorithms which determine the probability of observing a focal sequence, integrating over all possible paths through the data, under LS model [cite D-Palms]. This presents an opportunity to estimate parameters of interest while exploiting the tree-structure to gain log based speed-ups in exact their exact implementation.

@@

\renewcommand\thefigure{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand\thetable{S\arabic{table}}
\setcounter{table}{0}
\section*{Supplementary}
\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/supp_cophylogeny_long.pdf}
\caption{\label{fig:cophylogeny_long}Tanglegram equivalent to that in Figure~\ref{fig:cophylogeny}, but for the long ARG (i.e. subsampled to mid-2022). }
\end{figure}
